Clusterization Properties of D-optimal Designs in Infinite Dimensional Bayesian Linear Inverse Problems

### 2023-04-09 latex/intro.tex
Getting there

Designing experiments is usually done by optimizing some \emph{design
criterion}. This is true both for frequentists
\cite{silvey2013,Ucinski05} as well as for Bayesians
\cite{Chaloner1995} (see \cite{Chaloner1995} for an investigation of
the analogy between the two approaches). Among the plethora of design
criteria, \emph{D-optimal design} have one of the simplest and most
appealing motivation in the Bayesian context \cite{Chaloner1995}: a
D-optimal design maximizes the expected information gain (KL
divergence \cite{CoverThomas91}) between posterior and prior. For a
linear model in finite dimensions, with Gaussian prior and noise, a
D-optimal design minimizes the determinant of the posterior covariance
matrix. In a frequentist setting, a D-optimal design minimizes the
volume of the uncertainty ellipsoid \cite[page 16]{Ucinski05}, but
this is done for the Fisher information matrix and not the posterior
covariance. However, \cite{Chaloner1995} show that the latter is just
a regularized version of the former.

The previous discussion is classical for experimental design when
inference takes place over a finite (not too large) number of
parameters. The subject of optimal experimental design on function
spaces in a Bayesian context was pioneered by
\cite{AlexanderianGloorGhattas14, AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}. Similarly to the finite dimensional
case, a D-optimal design still maximizes the KL divergence between
posterior and prior. For a linear model with Gaussian prior and noise,
this amounts to minimizing the determinant of the posterior covariance
operator (understood as a product of its eigenvalues). Some
difficulties arise in the process, but remedies can be found as shown
in \cite{AlexanderianGloorGhattas14}.

It seems counter intuitive that when one computes an optimal design
using the D-optimality criterion, the optimization process results in
observations that are very similar. For example, if an observation is
thought of as measuring some function value at $\x \in \Omega
\subseteq \R^d, d=1,2,3$ (with added error) then the optimization
procedure sometimes places sensors in very close proximity to each
other (as can be seen in Figure \ref{fig:clusterization
  illustration}). Following \cite{Ucinski05}, we refer to this
phenomenon as \emph{sensor clusterization}.

\subsection{Related Work}
The phenomenon of sensor clusterization seems to be known in several
different contexts. In a frequentist and finite-dimensional context,
\cite{fedorov1996} and \cite[chapter 2.4.3]{Ucinski05} discuss this
phenomenon and suggest an approach called clusterization-free design.
In such designs, the user enforces observation locations to be far
from each other. One way to do this is by introducing correlated
errors which account for both observation error and model
error. Another method considered is imposing distance constraints
between observations. A somewhat different approach is suggested in
\cite[page 49]{fedorov2012}, where close observations are merged --- a
procedure which obviously does not avoid clusterization. The same
problem arises in time-series analysis for pharmacokinetic
experiments. The authors of \cite{hooker2009} suggest modeling
auto-correlation time in the noise model, which is equivalent to the
correlated errors mentioned above.

Any of the above mentioned approaches might serve as a remedy and push
sensors away from each other. Yet, none offers any insight as to why
clusterization occurs. Also, as better models are employed, model
error is decreased and the clusterization phenomenon will eventually
reappear. While these approaches are practical and help us avoid the
problem, they do not provide insight as to why sensors are clustering.

In the inverse problems community, work is mostly computational and
less theoretic. Model errors were considered in \cite{attia2020,
  koval2020}. The former study is focused on inferring a Quantity of
Interest (QoI). The focus of the latter is reducing forward solves,
using randomized linear algebra. Both studies present numerical
techniques for finding optimal designs when model error is
present. Both are restricted to linear inverse problems (although in
the latter the authors use their method on a nonlinear problem by
taking a Laplace approximation for the posterior). Both find an
optimal design by first solving a continuous problem for sensor
weights. Said solution is then sparsified to give a binary
design. Both studies are successful in the task of Bayesian
inversion. However, neither of these studies mention any effect model
errors can have on sensor clusterization. This study is mostly
theoretical and aims to fill the gap of understanding sensor
clusterization.

\subsection{Contribution}
We propose and study a relaxed and analytically tractable model for
D-optimal designs. Under this model, D-optimal designs are solutions
of a constrained optimization problem, formulated using Lagrange
multipliers. This allows me to rigorously show how model error
mitigates clusterization (Section \ref{section:non vanishing}). We
show that designs that exhibit clusterization are indeed optimal.

%% A beautiful mathematical structure arises in D-optimal designs

When no model error is present, the Lagrange multipliers problem boils
down to a nonlinear eigenvalue problem.
%% The operator for which eigenvectors and eigenvalues are sought is a
%% sum of two operators. The first is the prior covariance. The second
%% is an outer product of the observations (see section
%% \ref{section:vanishing} for details and exact statement).
This structure helps us to characterize D-optimal designs. The key
idea is that a D-optimal design reduces uncertainties where they are
highest first. This is the main conclusion of Theorem \ref{thm:char}
stated below.

@@

%% \begin{restatable*}[Flat posterior covariance]{corollary}{flatpost}\label{cor:equal eigs}
%%  %\begin{corollary}\label{cor:eigenvalues}
%%   If $\obs$ is D-optimal and $k := \rank \obs$, then:
%%   \begin{enumerate}
%%     \item The covariance of the pushforwad $\fwd_{*} \post$ is $\left ( (\fwd
%%       \prcov \fwd^*)^{-1} + \sigma^{-2} \obs^*\obs \right )^{-1}$.
%%     \item The eigenvalues of said operator are
%%       \begin{equation}
%%         \theta_i =
%%         \begin{cases}
%%           \left(\frac{1}{k} \sum_{j=1}^k \lambda_j^{-1} + \sigma^{-2}\frac{m}{k} \right )^{-1} & i \leq k \\
%%           \lambda_i &  i > k 
%%         \end{cases}
%%       \end{equation}
%%     \item The eigenvectors of said operator are the eigenvectors of
%%       $\fwd \prcov \fwd^*$
%%   \end{enumerate}
%% \end{restatable*}%\end{corollary}

%% \begin{restatable*}[Flat posterior covariance]{corollary}{flatpost}\label{cor:equal eigs}
%%   Consider a D-optimal design $\obs$ for a Bayesian linear inverse
%%   problem where $\fwd$ is the forward operator and $\pr$ is the
%%   prior. Then the eigenvectors of the covariance of the pushforward of
%%   the posterior $\fwd_{*} \post$ equal the eigenvectors of the
%%   covariance of the pushforward of the prior. Moreover, the
%%   eigenvalues are constant for eigenvectors $1 \leq i \leq k$ where $k
%%   =\rank \obs$.
%% \end{restatable*}

In the process, We generalize several lemmas from linear algebra to
infinite-dimensional settings. We prove a Matrix Determinant Lemma in
\ref{lemma:MDL}. We generalize a lemma due to Lax \cite{Lax07} for
calculating $\frac{\der}{\der t} \log \det (I + X(t))$, for an
operator valued function $X(t)$ in \ref{lemma:lax}. We show how to
decompose a symmetric positive definite matrix $m$ as $M = AA^t$,
where $A$ has unit norm columns in Lemma \ref{lemma:free}.

\subsection{Limitations}\label{subsec:limitations}
Two main drawbacks of the study presented here are: First, the relaxed
model does not consider any specific set of allowed
observations. Rather, we take observations in the unit ball in some
Hilbert space. This allows considerably less restrictive observations
than any real-life problem does. The second drawback is that we do not
show rigorously that clusterization necessarily occurs, we only show
that it is as reasonable as no clusterization.

>>>>

The problem of choosing the best set of measurements is known as the
problem of \emph{optimal experimental design}. Optimal designs are
found by optimizing some \emph{design criterion}. Many design criteria
exist, and in this study we focus on the Bayesian \emph{D-optimality
criterion}. The D-optimality criterion has a simple and appealing
motivation \cite{Chaloner1995}: a D-optimal design maximizes the
expected information gain (KL divergence \cite{CoverThomas91}) between
posterior and prior. For a linear model in finite dimensions, with
Gaussian prior and iid Gaussian noise, a D-optimal design minimizes
the determinant of the posterior covariance matrix\footnote{In a
frequentist setting, a D-optimal design minimizes the determinant of
the Fisher information matrix \cite[page 16]{Ucinski05}, so the
Bayesian criterion is just a regularized version of the frequentist
criterion \cite{Chaloner1995}.}.

Bayesian D-optimal designs for inverse problems on function spaces
were first studied by \cite{AlexanderianGloorGhattas14,
  AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}. Similarly to the finite dimensional
case, a D-optimal design maximizes the KL divergence between posterior
and prior measures. For a linear inverse problem with Gaussian prior
and iid Gaussian noise, a D-optimal design minimizes an expression
analogous to the determinant of the posterior covariance operator,
understood as a product of its eigenvalues.
%% Some difficulties arise in the process, but remedies can be found
%% as shown in \cite{AlexanderianGloorGhattas14}.

When one computes an optimal design using the D-optimality criterion,
the optimization process sometimes results in observations that are
very similar. For example, if an observation is thought of as
measuring some function value at $\x \in \Omega \subseteq \R^d,
d=1,2,3$ (with added error) then the optimization procedure sometimes
places sensors in very close proximity to each other (as can be seen
in Figure \ref{fig:clusterization illustration}). Following
\cite{Ucinski05}, we refer to this phenomenon as \emph{sensor
clusterization}.

Sensor clusterization is intuitively undesirable. Selecting a small
set of sensors may lead to incomplete or biased data, as certain parts
of the system or environment may be overlooked or
underrepresented. This can result in inaccurate or unreliable
estimates of the unknown parameters of interest, potentially leading
to incorrect conclusions or decisions.

%% \subsection{Related Work}
Several remedies are suggested in the literature to phenomenon of
sensor clusterization. In a frequentist and finite-dimensional
context, \cite{fedorov1996} and \cite[chapter 2.4.3]{Ucinski05}
discuss the clusterization phenomenon and suggest an approach called
clusterization-free design. In such designs, the user enforces
observation locations to be far from each other. One way to do this is
by introducing correlated errors which account for both observation
error and model error. Another method considered is imposing distance
constraints between observations. A somewhat different approach is
suggested in \cite[page 49]{fedorov2012}, where close observations are
merged --- a procedure which obviously does not avoid
clusterization. The same problem arises in time-series analysis for
pharmacokinetic experiments. The authors of \cite{hooker2009} suggest
modeling auto-correlation time in the noise model, which is equivalent
to the correlated errors mentioned above. Any of the above mentioned
approaches might serve as a remedy and push sensors away from each
other. Yet, none offers any insight as to why clusterization
occurs. Also, as better models are employed, model error is decreased
and the clusterization phenomenon will eventually reappear. While
these approaches are practical and help us avoid sensor
clusterization, they do not provide insight as to why sensors are
clustering.

Model errors were also considered for inverse problems in
\cite{attia2020, koval2020}. The former study is focused on inferring
a Quantity of Interest (QoI). The focus of the latter is reducing
forward solves, using randomized linear algebra. Both studies present
numerical techniques for finding optimal designs when model error is
present. Both are restricted to linear inverse problems (although in
the latter the authors use their method on a nonlinear problem by
taking a Laplace approximation for the posterior). Both find an
optimal design by first solving a continuous problem for sensor
weights. Said solution is then sparsified to give a binary
design. Both studies are successful in the task of Bayesian
inversion. Neither \cite{attia2020}, nor \cite{koval2020} mention any
effect model errors can have on sensor clusterization.

\subsection{Contribution}
The current study aims to fill the gap of understanding sensor
clusterization in infinte dimensional Bayesian linear inverse
problems. We propose and study a relaxed and analytically tractable
model for D-optimal designs. Under this model, D-optimal designs are
solutions of a constrained optimization problem, formulated using
Lagrange multipliers. The Lagrange multipliers formulation allows us
to rigorously show how model error mitigates clusterization (Section
\ref{section:non vanishing}), and why designs that exhibit
clusterization are indeed optimal when no model error is considered.

As a byproduct, the Lagrange multipliers problem for no model error
results in a characterization of D-optimal designs. The key idea is
that a D-optimal design reduces uncertainties where they are highest
first. This is the conclusion of part (5) of Theorem \ref{thm:char},
which we state below and prove in Section \ref{section:vanishing}:

@@

In the process of proving Theorem \ref{thm:char}, we prove a novel
lemma in linear algebra: We show how to decompose a symmetric positive
definite matrix $M$ as $M = AA^t$, where $A$ has unit norm columns
(Lemma \ref{lemma:lax}). We also generalize two lemmas from linear
algebra to infinite-dimensional settings: In Lemma \ref{lemma:MDL} we
prove a matrix determinant lemma in Hilbert spaces, and in Lemma
\ref{lemma:lax} we generalize a lemma \cite{Lax07} for calculating
$\frac{\der}{\der t} \log \det (I + X(t))$, where $X(t)$ is an
operator valued function.

\subsection{Limitations}\label{subsec:limitations}
The study presented here is not free from drawbacks. First, our
relaxed model does not correspond to any real-life problem. It is
generic enough to be analytically tractable, but one may argue it is
to far removed from any real application. To these claims I would
argue that Mathematics has a long history of studying models that are
bare-bones simplifications of real systems, e.g. the Ising model
\cite{cipra1987} the Lorenz system \cite{brin}, the SIR model
\cite{keeling2008}, to name a few. The second drawback of our model is
that we do not show rigorously that clusterization necessarily occurs,
we only show that it does not pose any obstruction to D-optimality.

### 2023-04-09 latex/paper.tex
Almost done?
Getting there

\title[Generic Properties of D-optimal Designs]{Generic Properties of D-optimal
  Designs in Infinite Dimensional Bayesian Linear Inverse Problems}

>>>>

\title[Clusterization Properties of D-optimal Designs]{Clusterization
  Properties of D-optimal Designs in Infinite Dimensional Bayesian
  Linear Inverse Problems}
